A Mechanised Library for Generic (Co)Programming

### 2024-01-25 paper/lensbans.tex
Recovering mutu, course-of-value and accu from hylomorphisms.

>>>>

\newcommand{\lbanana}{%
  \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
  \draw[join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;\mspace{1mu}}}
\newcommand{\rbanana}{%
  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
  \draw[join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;}}
% filled versions
\newcommand{\Lbanana}{%
  \mathopen{\mspace{1mu}\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
  \fill (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;\mspace{1mu}}}
\newcommand{\Rbanana}{%
  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
  \fill (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;}}
% % lens brackets (using TikZ)
% \newcommand{\llens}{%
%   \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
%   \draw[join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
%         to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;\mspace{1mu}}}
% \newcommand{\rlens}{%
%   \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
%   \draw[join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
%         to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;}}
% filled versions
\newcommand{\Llens}{%
  \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
  \fill (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
        to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;\mspace{1mu}}}
\newcommand{\Rlens}{%
  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
  \fill (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
        to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;}}

### 2024-01-25 paper/paper.tex
Recovering mutu, course-of-value and accu from hylomorphisms.

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}

\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}

\newcommand{\Bool}{\mathbb{B}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Nat}{\N}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

@@

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}

\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}

\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\InIso}{\text{In}}

\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}

@@

\maketitle
\section{Introduction}

Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

@@

\section{Background on Adjoint Folds}
Say that we want to prove that the two functions below are well-defined

@@

>>>>

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}
\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}
\newcommand{\Bool}{\mathbb{B}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Nat}{\N}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

@@

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\catafree}[1]{\Lbanana #1 \Rbanana}
\newcommand{\anacofree}[1]{\Llens #1 \Rlens}
\newcommand{\hylo}[2]{\cata{#1 \to #2}}
\newcommand{\cohylo}[2]{\ana{#1 \to #2}}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\head}{\operator{head}}
\newcommand{\tail}{\operator{tail}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}
\newcommand{\fmap}[1]{\text{fmap}\;#1}

\newcommand{\InOp}{\operator{in}^{\circ}}
\newcommand{\InIso}{\operator{in}}
\newcommand{\OutOp}{\operator{out}^{\circ}}
\newcommand{\OutIso}{\operator{out}}
\newcommand{\call}{\operator{call}}

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}

@@

\maketitle
\section{Introduction}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

@@

\section{Recursive Equations}

\subsection{Catamorphisms and Anamorphism}
A catamorphism is the unique program defined as
\[
  x = a \comp \fmap{x} \comp \InOp
\]
Can be implemented:
\[
  \cata{a} = \hylo{\InOp}{a}
\]

An anamorphism is the unique program satisfying the equation
\[
  x = \OutOp \comp \fmap{x} \comp{c}
\]
The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.

\paragraph{Free Monads}
Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
solution to the following domain equation
\[
  D^{*}A = A + DD^{*}A
\]
where $\eta_{A} : A \to D^{*}A$ is the left injection and
$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
$a : DA \to A$ there is a unique solution to the following recursive equation
\[
  x \comp \call_{A} = a \comp D(x)
\]
that is $\catafree{a} = \cata{[id_{A}; a]}$.

\paragraph{Cofree Comonads}
Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
greatest solution to the following domain equation
\[
  D_{\infty}A = A + DD_{\infty}A
\]
where $\epsilon_{A} : D_{\infty}A \to A$ is the first projection and
$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.

Given an algebra $c : A \to DA$ there is a unique solution to the following
recursive equation
\[
  \tail_{A} \comp x = D(x) \comp c
\]
that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$.

\subsection{Accumulators}
For an algebra $a : D(A^{P})\times P \to A$:
\begin{align*}
  & g : \mu D \times P \to B\\
  & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
\end{align*}

\paragraph{Hylo Implementation}
Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
solution to the following equation
\begin{align*}
  & f : \mu D \to A^{P}\\
  & f = a \comp \fmap (f) \comp \InOp
\end{align*}
thus $f = \hylo{\InOp}{a}$.

The complex recursion scheme is then
$g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.

\subsection{Mutu-Hylo}
For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
\begin{align*}
  & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
  & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
\end{align*}

\paragraph{Hylo Implementation}
Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
\begin{align*}
  & f : \mu D \to A_{1} \times A_{2}\\
  & f = a \comp \fmap{f} \comp \InOp
\end{align*}
for which the solution is an hylo $f = \hylo{\InOp}{a}$.

The mutumorphism can be implemented as
$(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.

\subsubsection{Example}
Say that we want to prove that the two functions below are well-defined

@@

\subsection{Course-of-Value Recursion}
First we have to assume a distributive law $\lambda : DG \to GD$.
For an algebra $a : D(D_{\infty} A) \to A$,
\begin{align*}
  & g : \mu F \to A\\
  & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}
where $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX  $ is a $D$-algebra on $GX$. The catamorphism of this algebra has the following type
\[
  \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
\]
which is the type of a $G$-algebra used to initialise the memory table.

\paragraph{Hylo Implementation}
Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
what is the unique solution of the following equation
\begin{align*}
  & f : \mu F \to G_{\infty} A\\
  & f = a \comp \fmap{f} \comp \InOp
\end{align*}
The solution to this equation can be found again using an hylo
$f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.
The course-of-value recursion can be recovered from this by setting
$g = \head \comp f$

\subsection{Course-of-Value Recursion with Finite Memory Table}
It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
any algebra $a$ generates a finite output.  In other words, as far as the
hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
\[
  G_{*} A = \mu X. A \times G X
\]
We now rewrite course-of-value recursion using the finite memory table
For an algebra $a : D(D_{*} A) \to A$,
\begin{align*}
  & g : \mu F \to A
  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}
